AAAI Press Formatting Instructions \\for Authors Using \LaTeX{} --- A Guide

### 2021-09-06 publications/iaai21/dora_outlier_detection_iaai.tex
Add to methods, including architecture diagram. #42

\subsection{DORA configurable pipeline} Steven please add details about the 
approach/design choices for this pipeline. Mention that additional data types, 
feature extraction methods, ranking algorithms, and results organization and 
visualization modules can easily be added and integrated if they follow the API. 
Hannah to add architecture figure. Add descriptions of each layer of modules: 
\paragraph{Data types} Hannah
\paragraph{Feature extraction} Hannah 
\paragraph{Ranking algorithms.} We implemented 7 algorithms for scoring and 
ranking samples by outlierness. [Describe each algorithm briefly and how they 
complement each other.]

The DEMUD algorithm~\citep{wagstaff:demud13} differs from other
outlier ranking methods: instead of independently scoring all
observations, DEMUD incrementally identifies the most unusual
remaining item, then incorporates it into the model of ``known''
(non-outlier) observations before selecting the next most unusual
item.  DEMUD's goal is to identify diverse outliers and avoid
redundant selections.  Once an outlier is encountered, repeated
occurrences of that outlier are deprioritized.  Methods that score
outliers independently maximize coverage of outliers, while DEMUD
maximizes the fast discovery of distinct outlier types.

\paragraph{Results organization} Eric, Umaa, Steven, Vinay
\paragraph{Outlier visualization} Jake?

@@

\subsubsection{Astrophysical objects in Dark Energy Survey}
Umaa

@@

\subsubsection{MNIST and FashionMNIST}
We added this to have a traditional benchmark and use case demonstration.

\subsection{Results}
Discuss the overall approach to the setup of the experiments. In each 
subsection discuss specific setup details (including evaluation metrics) for 
each dataset then the results from the experiments.

\subsubsection{Astrophysical objects in Dark Energy Survey}
Umaa

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/faw_combined_plot.pdf}
    \caption{Plot of number of true outliers ranked in top $N$ selections for 
    Earth science dataset.}
    \label{fig:faw_results}
\end{figure}

@@

\subsubsection{MNIST and FashionMNIST}
We added this to have a traditional benchmark and use case demonstration.

\section{Discussion}
\begin{itemize}
\item Discuss results from experiments
\item Challenges of evaluating results in outlier detection?
\item Downstream use of results in science applications/Path to deployment
\end{itemize}

\section{Conclusions}
Summarize conclusions and plans for future work, including public 
pip-installable package

[Note: these datasets could also be used by other researchers as OOD
 benchmark tasks?]

@@

\todo{Acknowledge DES?}
We thank the UN Food and Agriculture Organization (FAO) and Dr. Ritvik Sahajpal
(University of Maryland) for providing the fall army worm ground observation
dataset.

>>>>

\subsection{DORA configurable pipeline} 
Figure \ref{fig:dora} illustrates the architecture of DORA.
\todo{Steven please add a couple sentences about the
approach/design choices for this pipeline.} \todo{Describe the configuration
file setup. Discuss definitions of data to fit and data to score.
Maybe include example files in the appendix?}
We designed DORA to be readily extensible to support additional data types or 
formats, outlier detection algorithms, and results organization or 
visualization methods by writing new modules that follow the DORA API.

\paragraph{Data types.} We chose to initially implement data loaders for four
data types that are commonly used by the machine learning and domain science
communities: time series, feature vectors, images (grayscale or RGB), and 
$N$-band rasters. $N$-band rasters are essentially images in which every pixel
is associated with a location defined by a geographic coordinate reference 
system (e.g., latitude/longitude in degrees). Most satellite data are 
distributed as rasters; common formats include GeoTIFF, NetCDF, and 
\todo{HDF}. A data loader for each data type locates the data by the path(s)
defined in the configuration file and loads samples into a dictionary of numpy
 arrays indexed by the sample id. This $data_dict$ (\todo{format as 
 fixed-width}) is then passed to each of the ranking algorithms specified in 
 the configuration file.

\paragraph{Outlier ranking algorithms.} We implemented 7 algorithms for 
scoring and 
ranking samples by outlierness. We chose these algorithms to include a diverse
set of approaches to scoring outliers since different algorithms may have 
better performance for different datasets and use cases. 
%We describe each
%approach to scoring outliers and the associated methods below.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/dora-system-diagram-v5.pdf}
    \caption{DORA pipeline architecture.}
    \label{fig:dora}
\end{figure}

\subparagraph{Reconstruction error.}
\todo{Describe PCA: Hannah.}
The DEMUD algorithm~\citep{wagstaff:demud13} differs from other
outlier ranking methods: instead of independently scoring all
observations, DEMUD incrementally identifies the most unusual
remaining item, then incorporates it into the model of ``known''
(non-outlier) observations before selecting the next most unusual
item.  DEMUD's goal is to identify diverse outliers and avoid
redundant selections.  Once an outlier is encountered, repeated
occurrences of that outlier are deprioritized.  Methods that score
outliers independently maximize coverage of outliers, while DEMUD
maximizes the fast discovery of distinct outlier types.

\subparagraph{Distance.}
\todo{Describe RX and Local RX: Hannah.}

\subparagraph{Sparsity.}
\todo{Describe isolation forest: Umaa.}

\subparagraph{Likelihood.}
\todo{Describe PAE: Hannah or Bryce (if it is ready to be included.)}
\todo{Describe negative sampling: Steven.}

\paragraph{Results organization.} 
Each of the outlier ranking algorithms returns an array containing the sample
index, outlier score, and selection index (index of each sample after sorting
by outlier score). DORA provides results organization modules intended to help
 users interpret and make decisions based on these outputs. The simplest 
 module saves a CSV of the samples sorted by their outlier score 
 (i.e., selection order). Clustering the top
  $N$ outlier selections can enable users to investigate the different types of 
 outliers that might be present in the dataset; this could be especially useful
 for separating outliers caused by noise or data artifacts vs. scientifically 
 interesting samples. We implemented the K-means and self-organizing maps 
 (SOMs) algorithms for clustering the top-N outliers. For use cases in which an
 evaluation dataset containing known outliers is available, we provide a module
 to assess how well algorithm selections correlate with known outliers. This is
 done by plotting the number of true outliers vs. number of selections made. 
 This can also be represented as the precision at $N$ (\todo{add ref?}), i.e., 
 the number of true outliers divided by the number of selections $N$. Finally,
 we provide a module for plotting histograms of outlier scores to visualize the
 distribution of scores in the dataset (which may be, e.g., multimodal or 
 long-tailed).

\paragraph{Results visualization.} 
\todo{Jake: describe website if ready by deadline.}
\todo{Eric: describe causal graphs if added by deadline.}

@@

\subsubsection{Astrophysical objects in Dark Energy Survey.}
\todo{Umaa}

@@

\subsubsection{MNIST and FashionMNIST.}
We added this to demonstrate the capabilities of DORA with a traditional
benchmark dataset. \todo{Bryce. TBD if we will end up including due to
space and/or relevance.}

\subsection{Results}
In this section we discuss the experimental setup and results for each dataset.
We designed the experiments to be relevant to the real-world scientific use 
cases associated with each dataset domain.

\subsubsection{Astrophysical objects in Dark Energy Survey.}
\todo{Umaa}

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/faw_combined_plot.pdf}
    \caption{Number of true outliers ranked in top $N$ selections for 
    Earth science dataset.}
    \label{fig:faw_results}
\end{figure}

@@

\subsubsection{MNIST and FashionMNIST.}
\todo{Bryce. TBD if we will end up including due to
space and/or relevance.}

\section{Discussion}
\todo{ 
Potential topics TBD based on results:
\begin{itemize}
\item Discuss results from experiments
\item Challenges of evaluating results in outlier detection?
\item Downstream use of results in science applications
\end{itemize}
}
\paragraph{Path to deployment.} \todo{Hannah - describe pip-installable
package, open-source for extension by the wider scientific community, etc.}

\section{Conclusions}
\todo{
Summarize conclusions and plans for future work
[Note that these datasets could also be used by other researchers as OOD
 benchmark tasks?]
}

@@

\todo{Acknowledge DES?}
We thank the UN Food and Agriculture Organization (FAO) and Dr. Ritvik Sahajpal
(University of Maryland/NASA Harvest) for providing the fall army worm ground
 observation dataset.
